---
title: "Research Bazaar 2017"
author: "Leo Hardtke & Andrew Ferris"
date: "4 July 2017"
output:
  html_document:
    toc: true
    toc_float:
      collapsed: false
    theme: paper
runtime: shiny
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, cache=TRUE)
library(ggplot2)
library(shapefiles)
require(rgdal)
library(ggmap)
library(ggthemes)
library(broom)
library(shiny)
# read dataset and remove na's.
rainfall_raw <- read.csv("datasets/rainfall_bom_data.csv")
rainfall <- na.omit(rainfall_raw)
rainfall$date <- as.Date(rainfall$date)
```


## Interactive plotting

There are two main interactive plots types in R:

- Those were you can adjust the zoom level and pan (plotly) and
- those were you can change the parameters underlying your analysis and see the results immediately (Shiny)

### plotly
We will start with plotly, since it is easier to understand. Plotly is actually a JavaScript graphing library which was recently open sourced and has 'bindings' for several programming languages including R. It is specialized in creating interactive web-based graphs. Plotly graphs are rendered locally in R through the htmlwidgets framework.
Using the plot_ly function we could replicate the rainfall boxplot we did with ggplot as follows.

```{r}
library(plotly)
plot_ly(rainfall, y = ~rainfall, color = ~name, type="box")
```
Fancy! Isn't it? As you can see, plotly graphs are interactive. Click on legend entries to toggle traces, click-and-drag on the chart to zoom, double-click to autoscale, shift-and-drag to pan.

Fortunately we don't need to learn a new syntax thanks to the `ggplotly()` magic, we only need to store the `ggplot` object created by the `ggplot()` function.

```{r}
g <- ggplot(data=rainfall) +
  geom_line(aes(x=date, y=rainfall, color=name), show.legend = F) +
  facet_wrap(~name) +
   scale_x_date(date_breaks = "3 months", date_minor_breaks = "1 month", date_labels = "%b-%y", limits =as.Date(c("2016-01-01","2017-01-01"))) +
  scale_y_continuous("Rainfall [mm]", limits = c(0,300)) +
  theme_tufte()

ggplotly(g) 
```

90% accuracy! The only thing that didn't work was `show.legend = F`. The workaround is not very tricky... just add `%>% layout(showlegend = FALSE)` 

```{r}
g <- ggplot(data=rainfall) +
  geom_line(aes(x=date, y=rainfall, color=name), show.legend = F) +
  facet_wrap(~name) +
   scale_x_date(date_breaks = "3 months", date_minor_breaks = "1 month", date_labels = "%b-%y", limits =as.Date(c("2016-01-01","2017-01-01"))) +
  scale_y_continuous("Rainfall [mm]", limits = c(0,300)) +
  theme_tufte()
ggplotly(g) %>% layout(showlegend = FALSE)
```



### Shiny

Shiny creates more like little applications that you can either add to your interactive documents or embedded in a html document.

Lets see what we can do... 

```{r, cache=F}
sliderInput("binw", "Bin Width:", min = 1, max = 50, value = 30)

renderPlot({
  # draw the histogram with the specified number of bins
  ggplot(data=rainfall) + 
    geom_histogram(aes(x=rainfall), binwidth = input$binw)
})
```

Basically to create a Shiny app we need to define 2 part. The inputs, in this case a slider an then the plot that is wrapped in the `renderPlot()` function. From now on, your imagination is the only limit!